\documentclass[11pt]{article}

% Standard, non-exotic packages only (mirrors the C1 paper preamble).
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{mathtools}
\usepackage{booktabs}
\usepackage{array}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{colorlinks=true, linkcolor=blue, citecolor=blue, urlcolor=blue}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{proposition}{Proposition}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{remark}{Remark}

\newcommand{\R}{\mathbb{R}}
\newcommand{\Q}{\mathbb{Q}}
\newcommand{\chrom}{\chi}
\newcommand{\Myc}{M}

\title{Symmetry-broken SAT self-certifies the plane's
  $\chrom \ge 5$ lower bound\\(and where it stops)}

\author{Owen Kent\thanks{Independent researcher.
  Correspondence: \texttt{owenpkent@gmail.com}.}}
\date{June 2026}

\begin{document}
\maketitle

\begin{abstract}
The lower bound $\chrom(\R^2) \ge 5$ on the chromatic number of the plane rests
on certifying that specific finite unit-distance graphs (UDGs) are not
$4$-colorable. These certifications are $k$-colorability SAT instances whose
naive one-hot encoding carries the full $k!$ color-permutation symmetry, and
several of them have been reported as practically intractable. We report a simple
but decisive engineering result: exporting a sound symmetry break (clique-fixing
plus first-appearance color precedence) as a CNF preamble lets an off-the-shelf
CDCL solver decide all of them, each with an optional machine-checkable DRAT
proof. Concretely, $M^4(C_5)$ at $k=6$ (UNSAT, $22$ s), the $510$-vertex
Polymath16 graph at $k=4$ (UNSAT, $1.66$ s, previously abandoned as
SAT-intractable), and de Grey's $1585$-vertex graph at $k=4$ (UNSAT, $19.5$ min)
all become routine, where the naive encoding hits a wall. The entire known
$\chrom(\R^2) \ge 5$ computational program is thereby \emph{self-certifiable on a
single workstation}. We pair this with two methodological findings. First, a
\emph{portfolio law}: near the colorability phase boundary, single-solver runtime
is heavy-tailed and misleading; on one identical in-class instance CaDiCaL did not
finish in $12$ hours while MapleChrono returned SAT in $155$ seconds, a
$\ge\!280\times$ swing. Second, an honest \emph{regime boundary}: the symmetry
break crushes \emph{structured} UNSAT but does \emph{not} tame
\emph{phase-transition} hardness, so it certifies everything currently known and
does \emph{not} shortcut the search for a new $\chrom \ge 6$ object. The symmetry
break itself is standard; the contribution is wiring it into the
Hadwiger-Nelson pipeline, measuring that it crosses the specific walls, and
mapping where it stops.
\end{abstract}

\section{Introduction}
\label{sec:intro}

The Hadwiger-Nelson problem asks for the chromatic number $\chrom(\R^2)$ of the
unit-distance graph on the Euclidean plane. The classical bounds
$5 \le \chrom(\R^2) \le 7$ have stood for decades; the lower bound was raised from
$4$ to $5$ only recently, by de Grey~\cite{deGrey2018}, who exhibited a finite
unit-distance graph (UDG) on $1581$ vertices that is not $4$-colorable, verified
by SAT~\cite{Heule2018verify}. The Polymath16 project~\cite{Polymath16} and
Parts~\cite{Parts2020} subsequently reduced the order of such graphs to roughly
$510$ vertices. Every step of this lower-bound program reduces to the same
computational question:
\begin{quote}
\emph{Given a finite graph $G$ and an integer $k$, is $G$ $k$-colorable?}
\end{quote}
A $\chrom \ge 5$ certificate is a $4$-UNSAT verdict on a UDG; a reduction step is a
$5$-SAT witness on a smaller graph. The whole edifice is a stack of colorability
decisions.

These decisions are routinely handed to a black-box CDCL SAT solver through the
standard one-hot encoding (one Boolean $x_{v,c}$ per vertex-color pair). That
encoding has a structural defect for graph coloring: it carries the full
color-permutation symmetry, so every proper coloring appears $k!$ times in the
search space. A solver must either rediscover the symmetry or waste exponential
effort exploring symmetric subtrees. In practice this makes several of the
program's own instances awkward or intractable: the $510$-vertex graph at $k=4$
was at one point set aside as ``SAT-intractable,'' and $M^4(C_5)$ at $k=6$ (a
$95$-vertex triangle-free Mycielskian whose non-$6$-colorability certifies
$\chrom(M^4(C_5)) \ge 7$) was never completed by the naive route.

This note records that the obstruction is an encoding artifact, not a genuine
hardness, and that removing it is cheap. We describe a structure-first
colorability solver built for the program (Section~\ref{sec:solver}); the
sound symmetry break it searches natively, exported as a CNF preamble for a
production CDCL engine (Section~\ref{sec:export}); the resulting
\emph{self-certification} of the entire $\chrom(\R^2) \ge 5$ program on one machine
(Section~\ref{sec:results}); and two methodological findings that the program
paid for in compute and that we believe transfer to any UDG-colorability effort:
a portfolio law and a structured-versus-phase-transition regime boundary
(Section~\ref{sec:methodology}).

\paragraph{Contributions and honest scope.} (1) A from-scratch structure-first
colorability solver, validated against a production SAT backend with zero verdict
disagreements, that additionally counts colorings \emph{up to color permutation}
(a quantity SAT solvers do not expose). (2) A symmetry-broken CNF export that, on
the program's own instances, crosses two specific walls
($M^4(C_5)\,k{=}6$ and $P_{510}\,k{=}4$) and makes the de Grey lower bound
self-certifiable in $19.5$ minutes, each with an optional DRAT proof. (3) A
portfolio law and a regime boundary, stated as measured facts. We claim
\textbf{no} new bound on $\chrom(\R^2)$ and \textbf{no} novelty in the symmetry
break itself, which is the standard lex-leader / clique-fixing
construction~\cite{CrawfordSymmetry1996, SakallahSymmetry2009}. The contribution
is its application, the measurement that it crosses these particular walls, and
the honest delineation of where it does not help. Everything here is reproducible
on a single workstation with open-source solvers.

\section{Background}
\label{sec:background}

\paragraph{The one-hot encoding and its symmetry.} The standard CNF encoding of
$k$-colorability introduces a Boolean variable $x_{v,c}$ for each vertex
$v \in V(G)$ and color $c \in \{1,\dots,k\}$, with an at-least-one clause
$\bigvee_c x_{v,c}$ per vertex and an edge clause
$\overline{x_{u,c}} \vee \overline{x_{v,c}}$ for each edge $uv$ and color $c$
(at-most-one clauses per vertex are optional). Any permutation $\sigma \in S_k$ of
the colors maps a satisfying assignment to another satisfying assignment; the
$k!$ images are distinct points of the search space encoding the same coloring.
For UNSAT instances this multiplies the proof effort; for SAT instances it
scatters witnesses across symmetric frames.

\paragraph{Breaking the symmetry.} A standard remedy is a symmetry-breaking
predicate that keeps exactly one representative per orbit
\cite{CrawfordSymmetry1996, SakallahSymmetry2009}. For color symmetry two
ingredients suffice and are sound (every proper coloring retains exactly one
representative):
\begin{itemize}
  \item \textbf{Clique fixing.} A clique $Q = \{q_1,\dots,q_\omega\}$ must receive
        $\omega$ distinct colors; pin $q_i$ to color $i$ by unit clauses
        $x_{q_i,i}$. If $\omega > k$ this is an immediate UNSAT certificate.
  \item \textbf{First-appearance precedence.} Order the vertices $v_1,v_2,\dots$
        (clique first, then degree-descending) and require that color $c$ may
        first appear only after color $c-1$ has appeared:
        $x_{v_i,c} \rightarrow \bigvee_{j<i} x_{v_j,\,c-1}$, with rank-bound unit
        clauses forbidding $v_i$ from using colors above $i$. This is the
        lex-leader break specialized to color values.
\end{itemize}
The two together collapse the $k!$ relabelings to a single canonical coloring.

\paragraph{Checkable certificates.} A modern CDCL solver can emit a DRAT
proof~\cite{WetzlerHeuleHunt2014} of an UNSAT verdict, independently checkable by
a trusted tool. A symmetry-broken UNSAT proof certifies that the
\emph{symmetry-broken} CNF is unsatisfiable; since the added clauses are sound (an
orbit representative always survives), this entails that the original graph is not
$k$-colorable.

\section{A structure-first colorability solver}
\label{sec:solver}

We first built a colorability decision procedure from scratch, in pure Python,
specialized to the program's graphs (triangle-rich UDGs, Mycielskians, and
near-critical graphs). It is not a general SAT solver and is not meant to compete
with CDCL on industrial instances; it exists to make the program's structure
explicit and to serve as the source of the symmetry break exported in
Section~\ref{sec:export}.

\paragraph{Design.} Four ingredients, all from scratch with no SAT backend.
(i) \emph{Color-class symmetry breaking}: a vertex may introduce color $c$ only if
$c \le 1 + (\text{max color used so far})$, baking the canonical frame of
Section~\ref{sec:background} directly into the search. (ii) \emph{Clique seeding}:
a clique is colored $1..\omega$ up front, giving an instant $\omega > k$ UNSAT
certificate. (iii) \emph{MRV branching with forward checking}: branch on the
uncolored vertex with the fewest available colors, backtracking immediately on a
domain wipeout. (iv) \emph{Pattern counting}: the number of proper $k$-colorings
\emph{up to color permutation}, which SAT solvers do not provide and which
directly measures the ``liquidity'' of a graph's coloring space. A node budget
bounds every search so the solver never hangs.

\paragraph{Correctness by cross-validation.} The solver is validated, not
trusted. Across $84$ instance/$k$ pairs (a $60$-graph random self-test plus a
$24$-instance benchmark) its verdict agrees with a production SAT backend
(MapleChrono via PySAT~\cite{IgnatievPySAT}) on \emph{every} instance: zero
disagreements. Every returned coloring is checked proper, and the pattern counter
matches brute-force orbit counts on small graphs (e.g. $C_5$ has exactly $5$
three-coloring patterns).

\paragraph{The ceiling.} The solver has no clause learning, so on hard UNSAT
instances the search tree explodes: $882$ nodes at $M^2(C_5)$ but $447{,}720$ at
$M^3(C_5)$, and $M^4(C_5)\,k{=}6$ and $P_{510}\,k{=}4$ are out of reach in pure
Python within budget. Adding conflict-directed backjumping~\cite{Prosser1993} is a
large free win on hard structured instances (on $M^3(C_5)\,k{=}5$ UNSAT, $13.3$ s
$\to$ $1.2$ s, and a further $5\times$ under PyPy, $\sim\!78\times$ total), but
nogood learning without watched-literal propagation does not pay off in pure
Python, and $M^4(C_5)$ remains node-count-bound. The lesson is decisive: rather
than grow a pure-Python learning engine, \emph{export} the symmetry break to a
solver that already learns at full $1$UIP quality.

\section{Exporting the symmetry break to CDCL}
\label{sec:export}

The export emits the clique-fixing and first-appearance-precedence clauses of
Section~\ref{sec:background} as a preamble to the one-hot CNF, then hands the
result to a production CDCL solver (CaDiCaL~\cite{BiereCadical},
Glucose~\cite{AudemardSimon2009}, or MapleChrono, through
PySAT~\cite{IgnatievPySAT}). The solver inherits the color-symmetry break it
otherwise lacks, on top of its own learning, watched literals, VSIDS, and
restarts. A refinement pins an \emph{exact} maximum clique on the high-degree
core (cheap because the program's graphs are sparse, $K_4$-free, or
triangle-free) for a stronger immediate fix, falling back to a greedy clique.

\paragraph{Soundness and validation.} The added clauses are an orbit-selecting
predicate, so the symmetry-broken CNF is satisfiable iff $G$ is $k$-colorable. We
validated this empirically as well: over $1000+$ instances the symmetry-broken and
naive encodings returned identical verdicts with zero bad witnesses
(equisatisfiability check).

\section{Self-certifying the $\chrom(\R^2) \ge 5$ program}
\label{sec:results}

The export turns the program's hardest colorability decisions from walls into
seconds-to-minutes solves. All timings below are single isolated runs of
CaDiCaL~\cite{BiereCadical} on one workstation; ``wall'' marks an instance that
did not finish within the budget under the naive encoding.

\begin{table}[ht]
\centering
\caption{Walls crossed by the symmetry-broken export. The naive one-hot encoding
does not finish on the two middle rows; the symmetry-broken encoding decides them
in seconds.}
\label{tab:walls}
\begin{tabular}{l r r l}
\toprule
Instance & Naive one-hot & Symmetry-broken & Note \\
\midrule
$M^3(C_5)$ $k{=}5$ UNSAT & $7.0$ s & $\mathbf{0.05}$ s & $\sim\!146\times$ \\
$M^4(C_5)$ $k{=}6$ UNSAT & wall & $\mathbf{21.9}$ s & newly decided
  ($\chrom(M^4(C_5)) \ge 7$) \\
$P_{510}$ $k{=}4$ UNSAT & wall & $\mathbf{1.66}$ s & newly decided
  ($P_{510}$ genuinely $\chrom = 5$) \\
$P_{510}$ $k{=}5$ SAT & --- & instant & $5$-coloring witness \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{The whole lower-bound program, on one machine.} With the export, every
certificate the $\chrom(\R^2) \ge 5$ program depends on is reproducible on a
single workstation (Table~\ref{tab:battery}). In particular, de Grey's
$1585$-vertex graph is independently confirmed not $4$-colorable in $19.5$
minutes, a from-the-graph re-derivation of $\chrom(\R^2) \ge 5$, and all nine
$\chrom = 5$ lineage graphs are certified in seconds each.

\begin{table}[ht]
\centering
\caption{Self-certification battery (symmetry-broken encoding, CaDiCaL).}
\label{tab:battery}
\begin{tabular}{l r}
\toprule
Certificate & Time \\
\midrule
de Grey $1585$ $k{=}4$ UNSAT \quad ($\chrom(\R^2) \ge 5$) & $19.5$ min \\
nine lineage graphs ($510,517,529,553,610,633,803,826,874$) $k{=}4$ UNSAT &
  $1.5$--$9$ s each \\
$M^4(C_5)$ $k{=}6$ UNSAT \quad ($\chrom(M^4(C_5)) \ge 7$) & $22$ s \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Machine-checkable proofs.} Each symmetry-broken UNSAT solve optionally
emits a DRAT proof~\cite{WetzlerHeuleHunt2014}. The break also shrinks the proofs
dramatically: $M^3(C_5)\,k{=}5$ UNSAT is $13{,}263$ DRAT lines with the break
versus $879{,}467$ without (a $\sim\!66\times$ reduction), and $M^4(C_5)\,k{=}6$
UNSAT yields a complete $1{,}944{,}586$-line proof in $26$ s. We stress the honest
reading: the proof certifies that the \emph{symmetry-broken} CNF is unsatisfiable.
Because the symmetry clauses are sound, this entails non-colorability of the
graph; it is not a DRAT proof of the naive CNF, which is exactly the intractable
object the break avoids.

\section{Two methodological findings}
\label{sec:methodology}

The program paid for the following two facts in compute; we state them as measured
results because we expect them to transfer to any unit-distance colorability
effort.

\subsection{A portfolio law near the phase boundary}
\label{sec:portfolio}

While probing whether a $510$-vertex in-class UDG, augmented with several hundred
codegree-safe edges, becomes non-$5$-colorable, we hit a striking solver
disagreement on a single fixed instance (a $470$-edge augmentation, DIMACS
persisted):
\begin{quote}
\textbf{CaDiCaL did not decide it in $12{+}$ hours, while MapleChrono returned
SAT in $155$ seconds} --- a $\ge\!280\times$ heuristic-dependent swing on the
\emph{identical} CNF.
\end{quote}
This is the heavy-tailed runtime behavior expected of $k$-colorability instances
sitting at a satisfiability phase boundary, where the verdict is decided by a
handful of branching choices and different heuristics land on wildly different
trees. It retroactively explains an earlier instance abandoned as
``SAT-intractable'' after $66$ CaDiCaL-minutes: it was very likely tractable for a
different solver and was never re-tried. The operational law:

\begin{quote}
\textbf{Portfolio law.} Near-threshold $k$-colorability instances in the
unit-distance program must be run as a \emph{solver portfolio} (CaDiCaL for DRAT
certification; MapleChrono / Glucose for discovery), never a single solver. A
single-solver timeout is not evidence of hardness.
\end{quote}

\subsection{The regime boundary: structured UNSAT versus phase-transition hardness}
\label{sec:boundary}

It is tempting to read Section~\ref{sec:results} as ``symmetry breaking makes
these instances easy.'' That is true only for \emph{structured} UNSAT: sparse,
clearly-unsatisfiable graphs (the lineage, the Mycielski tower, de Grey's graph),
where the break removes the dominant redundancy and a learning solver finishes
fast. It does \emph{not} tame \emph{phase-transition} hardness. Concretely, the
``overshoot'' instances --- $P_{510}$ plus $\sim\!1000$ codegree-safe and
$K_4$-safe edges at $k=5$, deliberately pushed onto the $5$-SAT / $5$-UNSAT
boundary --- are a money pit for \emph{both} encodings: naive CaDiCaL did not
finish in $\sim\!100$ minutes, and the symmetry-broken encoding hit its conflict
budget at $500$k / $2$M / $10$M conflicts ($29$ s / $157$ s / $1387$ s,
undecided).

The consequence matters for the larger program. Symmetry breaking certifies
everything currently \emph{known} about $\chrom(\R^2) \ge 5$, but it does
\emph{not} unblock the search for a \emph{new} small $\chrom \ge 6$ object: that
obstruction is intrinsic to the colorability landscape, not an artifact of the
encoding. Distinguishing ``structured UNSAT'' from ``phase-transition hardness''
is the practical lesson; conflating them (as we briefly did) produces false
optimism about what a better encoding can buy.

\section{Reproducibility and honest limitations}
\label{sec:repro}

\paragraph{Reproducibility.} Everything above runs on a single workstation with
open-source components: the one-hot and symmetry-broken encoders, the structure-
first solver, and the PySAT~\cite{IgnatievPySAT} interface to
CaDiCaL~\cite{BiereCadical}, Glucose~\cite{AudemardSimon2009}, and MapleChrono.
The graphs are the published de Grey~\cite{deGrey2018} and
Polymath16~\cite{Polymath16, Parts2020} UDGs and the standard Mycielskians of
$C_5$~\cite{Mycielski1955}. UNSAT solves emit DRAT
proofs~\cite{WetzlerHeuleHunt2014} on request.

\paragraph{Limitations.}
\begin{enumerate}
  \item \emph{The symmetry break is standard.} Clique-fixing and lex-leader color
        precedence are textbook~\cite{CrawfordSymmetry1996, SakallahSymmetry2009}.
        We claim no novelty in the construction, only in its application to these
        instances and the measurement that it crosses these walls.
  \item \emph{Timings are single isolated runs}, not averaged benchmarks; the
        portfolio law of Section~\ref{sec:portfolio} is precisely why per-instance
        single-solver times must be read with care. The qualitative gap (wall
        versus seconds) is robust; the exact constants are not claimed to be.
  \item \emph{No bound on $\chrom(\R^2)$ moved.} This is tooling and
        certification, not a new mathematical result. The regime boundary of
        Section~\ref{sec:boundary} is explicit that the search for $\chrom \ge 6$
        is not advanced by faster decision procedures.
  \item \emph{The structure-first solver is a research instrument}, correct and
        transparent, not a competitor to CDCL at scale; its value is as the source
        of the exported break and as a pattern counter.
\end{enumerate}

\section{Conclusion}
\label{sec:conclusion}

The lower-bound side of the Hadwiger-Nelson problem is a stack of
$k$-colorability certificates, and the naive one-hot encoding handicaps every one
of them with the $k!$ color symmetry. Exporting a sound symmetry break to a
production CDCL solver removes the handicap and makes the \emph{entire} known
$\chrom(\R^2) \ge 5$ program self-certifiable on a single workstation, with
machine-checkable DRAT proofs, including instances previously reported
intractable. The same exercise yields two transferable lessons: run a portfolio
near the phase boundary, and do not mistake structured-UNSAT speed for progress on
phase-transition hardness. The break certifies what is known; finding the next
bound remains a mathematical problem, not a solver-engineering one.

\section*{Acknowledgments}
This note arose from an ongoing computational program on the Hadwiger-Nelson
problem. The graphs studied derive from the de Grey~\cite{deGrey2018} and
Polymath16~\cite{Polymath16} constructions. All computation was carried out on a
single workstation using open-source SAT solvers (CaDiCaL, MapleChrono, Glucose)
through the PySAT toolkit, with the pure-Python components run under PyPy. No
external funding supported this work.

\bibliographystyle{plain}
\bibliography{refs}

\end{document}
